---
title: "Test, Train, and Combined Data Sets"
author: "Whitney Holt"
date: "`r Sys.Date()`"
output: html_document
---

```{r include=FALSE, echo = FALSE, warnings = FALSE, message = FALSE}
library(tidyverse)
library(knitr)
library(readxl)
```

# Combined Data Set

## Loading & Joining Data

Loading in data sets:

```{r}
# Import data sets
transactional_data <- read.csv(file = "transactional_data.csv", stringsAsFactors = TRUE)
customer_profile <- read.csv(file = "customer_profile.csv", stringsAsFactors = TRUE)
customer_address_and_zip_mapping <- read.csv(file = "customer_address_and_zip_mapping.csv", stringsAsFactors = TRUE)
delivery_cost_data <- read_excel("delivery_cost_data.xlsx")
```

Combining transaction and customer data into one data set:

```{r}
# Join transactional and customer profile data
transaction_customer_data <- transactional_data %>%
                             left_join(customer_profile, by = c("CUSTOMER_NUMBER" = "CUSTOMER_NUMBER"))

# Join customer address data
transaction_customer_data <- transaction_customer_data %>%
                             left_join(customer_address_and_zip_mapping, by = c("ZIP_CODE" = "zip"))
```

## Feature Engineering from the EDA

Creating new variables from Exploratory Data Analysis (EDA):

**Total Units Ordered**

 - Description: used to quantify total units ordered, which are then used to determine when a threshold is reached
 - Calculation: sum of two existing variables

```{r}
# Total Units Ordered
transaction_customer_data$TOTAL_UNITS_ORDERED <- transaction_customer_data$ORDERED_GALLONS + transaction_customer_data$ORDERED_CASES
```

**Total Units Loaded**

 - Description: used to quantify total units loaded
 - Calculation: sum of two existing variables

```{r}
# Total Units Loaded
transaction_customer_data$TOTAL_UNITS_LOADED <- transaction_customer_data$LOADED_GALLONS + transaction_customer_data$LOADED_CASES
```

**Total Units Delivered**

 - Description: used to quantify total units delivered
 - Calculation: sum of two existing variables

```{r}
# Total Units Loaded
transaction_customer_data$TOTAL_UNITS_DELIVERED <- transaction_customer_data$DELIVERED_GALLONS + transaction_customer_data$DELIVERED_CASES
```

**Order Grouping**

 - Description: used to identify customers who are ordering less or more than 400 units per year
 - Calculation: logical statement utilizing level of detail (LOD) calculation functionality

```{r}
transaction_customer_data$ORDER_GROUPING <- (transaction_customer_data %>%
                                            group_by(YEAR, CUSTOMER_NUMBER) %>%
                                            mutate(TOTAL_UNITS_ORDERED = sum(TOTAL_UNITS_ORDERED)) %>%
                                            ungroup() %>%
                                            mutate(ORDER_GROUPING = as.factor(ifelse(TOTAL_UNITS_ORDERED < 400, 
                                                                                                'Customers Ordering Less than 400', 
                                                                                                'Customers Ordering More than 400'))))$ORDER_GROUPING
```

**Customer Group Number (Rollup)**

- Description: used to aggregate transactions up to the customer group number, if desired
- Calculation: logical statement using two existing variables

```{r}
transaction_customer_data$CUSTOMER_GROUP_NUMBER_ROLLUP <- ifelse(!is.na(transaction_customer_data$PRIMARY_GROUP_NUMBER), 
                                                                 transaction_customer_data$PRIMARY_GROUP_NUMBER, 
                                                                 transaction_customer_data$CUSTOMER_NUMBER)
```

## New Feature Engineering: Calculating Delivery Cost per Customer

Steps provided by SWIRE:

 1. Determine the volume range for each category (Fountain and Bottles/ Cans)
 2. Retrieve the median delivery cost per unit from the data set.
 3. Multiply the cost per unit by the quantity purchased for each category.
 4. Sum the delivery costs to get the total.
 
We'll start by calculating the total volume purchased by each customer in each category per year. 

```{r}
# Summarize the total volume purchased for each category
annual_volume <- transaction_customer_data %>%
  group_by(YEAR, CUSTOMER_NUMBER, COLD_DRINK_CHANNEL) %>%
  summarize(
    Total_Fountain = sum(ORDERED_GALLONS),
    Total_Bottles_Cans = sum(ORDERED_CASES)
  ) %>%
  ungroup()

# View the result
head(annual_volume, n = 5)
tail(annual_volume, n = 5)
```

Next, we need to determine the range each fountain and bottle/ can value falls into, writing a function to do so.

```{r}
# Ensure Vol Range is a character
delivery_costs <- delivery_cost_data %>%
  mutate(`Vol Range` = as.character(`Vol Range`))

# Function to get the delivery cost based on volume range
get_delivery_cost <- function(volume, category, channel) {
  if (volume >= 0 & volume <= 149) {
    cost <- delivery_costs %>%
      filter(`Cold Drink Channel` == channel, `Applicable To` == category, `Vol Range` == "0 - 149") %>%
      select(`Median Delivery Cost`) %>%
      pull()
  } else if (volume >= 150 & volume <= 299) {
    cost <- delivery_costs %>%
      filter(`Cold Drink Channel` == channel, `Applicable To` == category, `Vol Range` == "150 - 299") %>%
      select(`Median Delivery Cost`) %>%
      pull()
  } else if (volume >= 300 & volume <= 449) {
    cost <- delivery_costs %>%
      filter(`Cold Drink Channel` == channel, `Applicable To` == category, `Vol Range` == "300 - 449") %>%
      select(`Median Delivery Cost`) %>%
      pull()
  } else if (volume >= 450 & volume <= 599) {
    cost <- delivery_costs %>%
      filter(`Cold Drink Channel` == channel, `Applicable To` == category, `Vol Range` == "450 - 599") %>%
      select(`Median Delivery Cost`) %>%
      pull()
  } else if (volume >= 600 & volume <= 749) {
    cost <- delivery_costs %>%
      filter(`Cold Drink Channel` == channel, `Applicable To` == category, `Vol Range` == "600 - 749") %>%
      select(`Median Delivery Cost`) %>%
      pull()
  } else if (volume >= 750 & volume <= 899) {
    cost <- delivery_costs %>%
      filter(`Cold Drink Channel` == channel, `Applicable To` == category, `Vol Range` == "750 - 899") %>%
      select(`Median Delivery Cost`) %>%
      pull()
  } else if (volume >= 900 & volume <= 1049) {
    cost <- delivery_costs %>%
      filter(`Cold Drink Channel` == channel, `Applicable To` == category, `Vol Range` == "900 - 1049") %>%
      select(`Median Delivery Cost`) %>%
      pull()
  } else if (volume >= 1050 & volume <= 1199) {
    cost <- delivery_costs %>%
      filter(`Cold Drink Channel` == channel, `Applicable To` == category, `Vol Range` == "1050 - 1199") %>%
      select(`Median Delivery Cost`) %>%
      pull()
  } else if (volume >= 1200 & volume <= 1349) {
    cost <- delivery_costs %>%
      filter(`Cold Drink Channel` == channel, `Applicable To` == category, `Vol Range` == "1200 - 1349") %>%
      select(`Median Delivery Cost`) %>%
      pull()
  } else if (volume >= 1350) {
    cost <- delivery_costs %>%
      filter(`Cold Drink Channel` == channel, `Applicable To` == category, `Vol Range` == "1350+") %>%
      select(`Median Delivery Cost`) %>%
      pull()
  } else {
    cost <- 0
  }
  return(ifelse(length(cost) > 0, cost, 0))
}
```

Next, we'll calculate the total delivery cost for each customer.

```{r}
# Calculate the total delivery cost for each customer
annual_volume <- annual_volume %>%
  rowwise() %>%
  mutate(
    Fountain_Cost = get_delivery_cost(Total_Fountain, "Fountain", COLD_DRINK_CHANNEL) * Total_Fountain,
    Bottles_Cans_Cost = get_delivery_cost(Total_Bottles_Cans, "Bottles and Cans", COLD_DRINK_CHANNEL) * Total_Bottles_Cans,
    Total_Delivery_Cost = Fountain_Cost + Bottles_Cans_Cost
  ) %>%
  ungroup()

# View the result
head(annual_volume, n = 5)
tail(annual_volume, n = 5)
```

Last, we'll add the total delivery cost column to our main data set.

```{r}
# Merge the total delivery cost back into the transaction_customer_data table
transaction_customer_data <- transaction_customer_data %>%
  left_join(annual_volume %>% select(CUSTOMER_NUMBER, YEAR, Total_Delivery_Cost), 
            by = c("CUSTOMER_NUMBER" = "CUSTOMER_NUMBER", "YEAR" = "YEAR"))

# View the result
head(transaction_customer_data, n = 5)
tail(transaction_customer_data, n = 5)
```

### Exporting Data

```{r}
write.csv(transaction_customer_data, "transaction_customer_data.csv", row.names = FALSE)
```

# Test & Train Data Set

## Defining test and train sets of disaggregated data

Because some customers have data in just 2023 or 2024, we want to make sure we split those customers into train and test sets proportionately.

The plan is to:

 - Identify customers with data in 2023, 2024, both years, and only one of the years.
 - Split the customers into train and test sets based on the specified proportions.
 - Create the train and test sets using the transaction_customer_data table.
 
```{r}
# Identify customers with data in 2023 and 2024
customers_2023 <- transaction_customer_data %>%
  filter(YEAR == 2023) %>%
  select(CUSTOMER_NUMBER) %>%
  distinct()

customers_2024 <- transaction_customer_data %>%
  filter(YEAR == 2024) %>%
  select(CUSTOMER_NUMBER) %>%
  distinct()

# Identify customers with data in both years, only 2023, and only 2024
customers_both_years <- intersect(customers_2023$CUSTOMER_NUMBER, customers_2024$CUSTOMER_NUMBER)
customers_only_2023 <- setdiff(customers_2023$CUSTOMER_NUMBER, customers_2024$CUSTOMER_NUMBER)
customers_only_2024 <- setdiff(customers_2024$CUSTOMER_NUMBER, customers_2023$CUSTOMER_NUMBER)

# Function to split customers into train and test sets
split_customers <- function(customers, train_ratio = 0.7) {
  set.seed(123) # For reproducibility
  train_indices <- sample(seq_len(length(customers)), size = floor(train_ratio * length(customers)))
  train_customers <- customers[train_indices]
  test_customers <- customers[-train_indices]
  return(list(train = train_customers, test = test_customers))
}

# Split customers into train and test sets
split_both_years <- split_customers(customers_both_years)
split_only_2023 <- split_customers(customers_only_2023)
split_only_2024 <- split_customers(customers_only_2024)

# Combine train and test customers
train_customers <- c(split_both_years$train, split_only_2023$train, split_only_2024$train)
test_customers <- c(split_both_years$test, split_only_2023$test, split_only_2024$test)

# Create train and test sets
train_set <- transaction_customer_data %>%
  filter(CUSTOMER_NUMBER %in% train_customers)

test_set <- transaction_customer_data %>%
  filter(CUSTOMER_NUMBER %in% test_customers)

# View the results
print("Train Set:")
head(train_set, n = 5)
tail(train_set, n = 5)
print("Test Set:")
head(test_set, n = 5)
tail(test_set, n = 5)
```

Let's test to make sure each customer (and all of their transactions) end up in either the train set or the test set, and aren't split among the two:

```{r}
# Check for overlap
overlap_customers <- intersect(train_customers, test_customers)
if (length(overlap_customers) == 0) {
  print("No overlap between train and test customers.")
} else {
  print("Overlap found between train and test customers.")
}
```

### Exporting Data

```{r}
write.csv(train_set, "train_set.csv", row.names = FALSE)
write.csv(test_set, "test_set.csv", row.names = FALSE)
```

## Defining test and train sets of aggregated data

These sets are still quite large let's aggregate each row up to the customer and year level so there is one row per customer per year. This will decrease the time required to create and test the models.

```{r}
# Assigning columns a group
dimensions <- c("ORDER_TYPE", 
                "PRIMARY_GROUP_NUMBER",
                "FREQUENT_ORDER_TYPE",
                "FIRST_DELIVERY_DATE",
                "ON_BOARDING_DATE",
                "COLD_DRINK_CHANNEL",
                "TRADE_CHANNEL",
                "SUB_TRADE_CHANNEL",
                "LOCAL_MARKET_PARTNER",
                "CO2_CUSTOMER",
                "ZIP_CODE",
                "full.address",
                "ORDER_GROUPING",
                "CUSTOMER_GROUP_NUMBER_ROLLUP")

measures <- c("ORDERED_CASES", 
              "LOADED_CASES", 
              "DELIVERED_CASES",
              "ORDERED_GALLONS",
              "LOADED_GALLONS",
              "DELIVERED_GALLONS",
              "TOTAL_UNITS_ORDERED",
              "TOTAL_UNITS_LOADED",
              "TOTAL_UNITS_DELIVERED") 

drop_columns <- c("WEEK", 
                  "TRANSACTION_DATE") 

already_aggregated <- "Total_Delivery_Cost" 

# Aggregate the data
aggregated_data <- transaction_customer_data %>%
  select(-one_of(drop_columns)) %>% # Drop unnecessary columns
  group_by(CUSTOMER_NUMBER, YEAR) %>%
  summarize(across(all_of(dimensions), first), # Keep dimensions as-is
            across(all_of(measures), sum, .names = "sum_{col}"), # Sum measures
            across(all_of(already_aggregated), first)) # Keep already aggregated column as-is

# View the results
print("Aggregated Data:")
head(aggregated_data, n = 5)
tail(aggregated_data, n = 5)

# Aggregate the train set
aggregated_train_set <- train_set %>%
  select(-one_of(drop_columns)) %>% # Drop unnecessary columns
  group_by(CUSTOMER_NUMBER, YEAR) %>%
  summarize(across(all_of(dimensions), first), # Keep dimensions as-is
            across(all_of(measures), sum, .names = "sum_{col}"), # Sum measures
            across(all_of(already_aggregated), first)) # Keep already aggregated column as-is

# Aggregate the test set
aggregated_test_set <- test_set %>%
  select(-one_of(drop_columns)) %>% # Drop unnecessary columns
  group_by(CUSTOMER_NUMBER, YEAR) %>%
  summarize(across(all_of(dimensions), first), # Keep dimensions as-is
            across(all_of(measures), sum, .names = "sum_{col}"), # Sum measures
            across(all_of(already_aggregated), first)) # Keep already aggregated column as-is

# View the results
print("Aggregated Train Set:")
head(aggregated_train_set, n = 5)
tail(aggregated_train_set, n = 5)

print("Aggregated Test Set:")
head(aggregated_test_set, n = 5)
tail(aggregated_test_set, n = 5)
```

Lastly, I want to make sure that I am using 2023 data to predict Total Units ordered in 2024.

```{r}
# Separate the data into 2023 and 2024 subsets
train_data_2023 <- aggregated_train_set %>% filter(YEAR == 2023)
train_data_2024 <- aggregated_train_set %>% filter(YEAR == 2024)
test_data_2023 <- aggregated_test_set %>% filter(YEAR == 2023)
test_data_2024 <- aggregated_test_set %>% filter(YEAR == 2024)

# Merge the 2023 and 2024 data based on customer number
train_merged_data <- merge(train_data_2023, train_data_2024, by = "CUSTOMER_NUMBER", suffixes = c("_2023", "_2024"))
test_merged_data <- merge(test_data_2023, test_data_2024, by = "CUSTOMER_NUMBER", suffixes = c("_2023", "_2024"))

# Select relevant
train_set <- train_merged_data %>% select(CUSTOMER_NUMBER, ends_with("_2023"), sum_TOTAL_UNITS_ORDERED_2024)
test_set <- test_merged_data %>% select(CUSTOMER_NUMBER, ends_with("_2023"), sum_TOTAL_UNITS_ORDERED_2024)
```

### Exporting Data

```{r}
write.csv(train_set, "train_set_aggregated.csv", row.names = FALSE)
write.csv(test_set, "test_set_aggregated.csv", row.names = FALSE)
```




















